---
title: "SALURBAL DBT Censorship v2"
subtitle: "This notebook is use to develop and analyze the censorship process for SALURBAL data v2 API"
author: 
  - name: Ran Li (Maintainer)
    orcid: 0000-0002-4699-4755
  - name: Jessica Uruchima (Context Partner)
date: last-modified
format: 
  html:
    toc: true
    self-contained: true
    code-fold: false
    df-print: kable
    code-tools: true
    code-block-border-left: '#39b69d'
    code-block-bg: '#ebf0ef'
comments:
  hypothesis: 
    theme: clean
editor: source
execute:  
  warning: false
  message: false
  eval: true
editor_options: 
  chunk_output_type: console
---


```{r}
#| code-summary: 'Setup'
#| code-fold: true

setwd(file.path(here::here(),"code/dbt/.automations"))

pacman::p_load(tidyverse, arrow, reactable, assertr, glue)

## Local variables
dbt = '//files.drexel.edu/colleges/SOPH/Shared/UHC/Projects/Wellcome_Trust/Data Methods Core/Dashboards/dbt/salurbal-dbt-server/models/stage/'
```

# 1. Intro


This was a process used to add one last layer of censroship to SALURBAL data. The first time implement is documented in [this GitHub issue](https://github.com/Drexel-UHC/salurbal-etl/issues/124); Let's get the most recent censorship file. 

```{r}

## Get most recent censorship date
most_recent_date = list.files('censorship/') %>% 
   keep(~ str_detect(.x, '^202[0-9]-[0-9]{2}-[0-9]{2}$')) %>% 
   parse_date() %>%
   max() %>% 
  as.character()


## Import 
df_censor_previous = glue("censorship/{most_recent_date}/variable_censorship_{str_remove_all(most_recent_date,'-')}.csv")|> 
  read_csv()

## Preview
glimpse(df_censor_previous)
```


There are six variables.

- 2 composite keys
   - var_name: variable identifier
   - dataset_id: dataset identifer
- 3 metadata fields to guide the censorship status input
   - var_label: short name for variable
   - var_def: variable defintion
   - domain
   - subdomain
   - public_any: if any is public
- 1 empty column for data managers to fill out:
    - keep: this is something to fill out by DMC. If the value is `1` then the L1AD and most recent year of data cane be released on the portal. If NA or 0 then the varaible cannot be released on the data portal for download. 
    
    
    
#  2. Components

We can generate this template from with our pre-DMC censored API which is contained within a DBT model called `int__precensored_admin_cube.parquet`. Let's connect to taht first.

```{r}
precensor_api = file.path(dbt,'int__precensored_admin_cube.parquet') |> 
  open_dataset()
```

There are two different componetns:

1. Composite keys (dataset and variable identifiers)
2. Summaries of percent of datapoints that are public

## Composite Keys

Let's get the composite keys.

```{r}

df_composite_keys = precensor_api |> 
  select(dataset_id, dataset_version, dataset_instance, 
         observation_type, domain, subdomain,
         var_name,var_label,var_def) |> 
  distinct() |> 
  collect() 
```


## Percent Public

The original `public` variable was extracted from DUA's. However there are many layers of additional censorship before data gets into the API. Some of the censorship is done in the  `int__precensored_admin_cube` dbt model which is used to generate the DMC censorship templates. The logic is as follows:

```
SELECT 
    *,
    IF (observation_type NOT IN ('area-level'), '1', '0') as censor_obs_type,
    IF (geo NOT IN ('L1AD'), '1', '0') as censor_geo,
    IF (time_resolution_type NOT IN ('year', 'year range'), '1', '0') as censor_time_resolution,
    IF (estimate_type  NOT IN ('estimate'), '1', '0') as censor_estimate_type,
    CASE
        WHEN censor_obs_type = '1'         THEN '88_obs_type'
        WHEN censor_geo = '1'              THEN '88_geo'
        WHEN censor_time_resolution = '1'  THEN '88_time_resolution'
        WHEN censor_estimate_type = '1'    THEN '88_estimate_type'
        WHEN censor_most_recent_year = '1' THEN '88_not_recent_year'
        ELSE public
    END AS public__pre_dmc_censor
```
 
Here we use this to calculate summaries of what percent of data points per variable is still public after these pre-DMC censorship steps.

```{r}

df_public  = precensor_api |> 
  count(dataset_instance, var_name,public, public__pre_dmc_censor) |> 
  collect() |> 
  verify(not_na(public__pre_dmc_censor)) 

df_pct_public = df_public|> 
  mutate(public_true = ifelse(public__pre_dmc_censor == 1,
                              'n_datapoints_public_1',
                              'n_datapoints_public_0')) |> 
  select(-public__pre_dmc_censor) |> 
  summarize(n = sum(n), .by = c(dataset_instance, var_name, public_true)) |> 
  pivot_wider(names_from = public_true, values_from = n) |> 
  mutate(across(everything(), ~replace_na(., 0))) |> 
  mutate(n_datapoints = n_datapoints_public_1 + n_datapoints_public_0,
         pct_public = round((n_datapoints_public_1/n_datapoints) * 100,2)) |> 
  left_join(df_composite_keys |> 
              select(dataset_instance,var_name, observation_type, domain) |> 
              distinct())

```

# 3. Validation

Let's do some analysis to better understand the bigger picture and to do validations on these components before we compile!

## Variables

Visualize distribution.

```{r}
#| code-fold: true
#| code-summary: "Visualize distirbution of variable counts"

custom_colors <- c("area-level" = "#3b7dd8", "record-level" = "#d84d3b")

ggplot(df_composite_keys, aes(x = domain, fill = observation_type)) +
  geom_bar() +
  coord_flip() +
  scale_fill_manual(values = custom_colors) +
  labs(title = "Variable Count Distribution \nAcross Domains/Observation-types",
       x = "Domain",
       y = "Number of Variables",
       fill = "Observation Type") +
  theme_minimal() +
  theme(legend.position = "bottom")
  
```

Here is a table for details

```{r}
#| code-fold: true

df_composite_keys |> 
  count(observation_type, dataset_instance, name = 'n_variables') |> 
  arrange(desc(n_variables)) |> 
  reactable(searchable = T)
```
 
## Percent Public


Let's do a analysis on distribution of pct_public at variable level.

```{r}
#| code-fold: true
#| code-summary: "Visualize distirbution of variable pct_public"

df_pct_public |> 
  ggplot(aes(x=observation_type, y = pct_public)) +
  geom_boxplot() +
  facet_wrap(~domain)+
  theme_bw() +
  labs(
    title = "Variable pct_public distribution by domain and observation type",
  )
  
```

Here is the table for details.

```{r}
#| code-fold: true

df_pct_public |> 
  select(-domain) |> 
  select(observation_type, everything()) |> 
  arrange(desc(pct_public)) |> 
  reactable()
```


## Data point censorship


```{r}
#| code-fold: true
#| code-summary: "Visualize distirbution of data point censorship type"

df_plot_data = df_public |> 
  left_join(df_composite_keys) |> 
  summarize(
    n = sum(n),
    .by = c(observation_type, domain, public__pre_dmc_censor)) 

ggplot(data = df_plot_data, 
       aes(x = n, y = domain, fill = public__pre_dmc_censor)) +
  geom_col() +
  labs(title = "Distribution of data point Censorship Status by Domain",
       x = "# Data points", 
       y = "Domain",
       fill = "Censorship Status") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set3") +
  facet_wrap(~observation_type) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_continuous(labels = scales::comma)

```


A few interesting stories here:

- All recorded level are censored due to Observation Type (PINK)
- Physical natural envrionemnt  (aka BEC)
   - Most censured due to `88_geo` which limits to only L1AD (this datasets has a lot L2/L2_5/L3 datapoints (ORANGE)
   - also a lot of `88_time_resolution` censoring because this dataset has a lot of longitudinal data. But we only share cross sectional data. (GREY)
- Mortality data most censored because a lot of it is 'iteration' or 'sampled' data and are censored because theya re not vlaid `88_estimate_types` e.g. `estimates`. (BLUE)
- SEC data is censored mostly because its not L1AD. (ORANGE)

  
## Assertions tests


Let's do some assertion testing specific to this release:

- record-level data is not shared (`pct_public == 0`)

```{r}

valid = df_pct_public %>%
  filter(observation_type == "record-level") |> 
  verify(pct_public == 0)
```
 

# 4. Template for DMC

## Compile

Let's compile and export. Note we also include previous censorship values for reference:

- keep_v1: the `keep` values used in the v1.0 release

 
```{r}
df_censorship_template = df_composite_keys |> 
  left_join(df_pct_public) |> 
  left_join(df_censor_previous |> select(dataset_instance, var_name, keep_previous = keep)) |>  
  mutate(keep_previous = as.character(keep_previous) |> 
           replace_na("")) |> 
  verify(not_na(pct_public)) |> 
  verify(between(pct_public,0, 100)) |>
  mutate(keep = '') |> 
  select(observation_type, domain, 
         dataset_instance, var_name, 
         var_label,var_def, pct_public, keep_previous, keep) |> 
  arrange(observation_type,dataset_instance, var_name)

```

## Validate

Let's do a few validations particularly comparing with previous censorship templates to make sure that no variables were dropped. Let's see which variables are missing.

```{r}
df_dropped_variables = df_censor_previous |> 
  mutate(dataset_id = dataset_instance |> str_remove_all("_v1.0|_v2.0|_v3.0")) |> 
  select(dataset_id, var_name) |> 
  mutate(old = '1') |> 
  full_join(select(df_composite_keys,dataset_id, var_name) |> mutate(new = '1')) |> 
  filter(is.na(new)) %>%
  verify(nrow(.)==0)
```


Confirmed that the new template has all the previous variables. 

## Preview

```{r}
df_censorship_template |> 
  select(observation_type, dataset_instance, var_name, pct_public, contains('keep')) |>
  reactable()
```

## Export 

```{r}
df_censorship_template |> 
  write_csv(glue("censorship/censorship_{Sys.Date() |> str_remove_all('-')}_template.csv"))
```

# 5. DBT Validation

## Admin cube

Just to check `on_portal` is valid.

```{r}

core = file.path(dbt,'core__admin_cube.parquet') |> 
  open_dataset()

dfa = core |> 
  count(public,
        public__pre_dmc_censor,
        dmc_keep_var,
        on_portal) |> 
  collect() |> 
  mutate(
    invalid = any(
        dmc_keep_var != '1' & on_portal == '1',
        public__pre_dmc_censor != '1'  & on_portal == '1',
        public != '1' & on_portal == '1'
      )
  ) |> 
  verify(!invalid) |> 
  arrange(on_portal)


dfa |> reactable(searchable = T)
```